---
title: 'Homework #4 STA 360'
author: "Jerry Xin STA 360: Homework 4"
date: "Due Friday September 17th, 5 PM EDT"
output:
  pdf_document: 
    fig_height: 4
    fig_width: 6
---

```{r setup, echo = F}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.align = "center")
```

```{r load-packages, warning = F, message = F}
library(plyr)
library(ggplot2)
library(dplyr)
library(xtable)
library(reshape)
library(tidyverse)
```

1. (10 points, 5 points each) Hoff, 3.10 (Change of variables).


2. Lab component (25 points total) Please refer to lab 4 and complete tasks
4—5.


# (a) (10) Task 4 (Finish for Homework)

```{r}
set.seed(123)
# input data
# spurters
x = c(18, 40, 15, 17, 20, 44, 38)
# control group
y = c(-4, 0, -19, 24, 19, 10, 5, 10,
      29, 13, -9, -8, 20, -1, 12, 21,
      -7, 14, 13, 20, 11, 16, 15, 27,
      23, 36, -33, 34, 13, 11, -19, 21,
      6, 25, 30,22, -28, 15, 26, -1, -2,
      43, 23, 22, 25, 16, 10, 29)
# store data in data frame 
iqData = data.frame(Treatment = c(rep("Spurters", length(x)), 
                                  rep("Controls", length(y))),
                                  Gain = c(x, y))
```

```{r}
xLimits = seq(min(iqData$Gain) - (min(iqData$Gain) %% 5),
              max(iqData$Gain) + (max(iqData$Gain) %% 5),
              by = 5)

ggplot(data = iqData, aes(x = Gain, fill = Treatment, colour = I("black"))) + 
  geom_histogram(position = "dodge", alpha = 0.5, breaks = xLimits, closed = "left")+
  scale_x_continuous(breaks = xLimits, 
                     expand = c(0,0))+ 
  scale_y_continuous(expand = c(0,0), 
breaks = seq(0, 10, by = 1))+
  ggtitle("Histogram of Change in IQ Scores") + labs(x = "Change in IQ Score", 
                                                     fill = "Group") + 
                                               theme(plot.title = element_text(hjust = 0.5))  
```
```{r}
prior = data.frame(m = 0, c = 1, a = 0.5, b = 50)
findParam = function(prior, data){
  postParam = NULL
  c = prior$c
  m = prior$m
  a = prior$a
  b = prior$b
  n = length(data)
  postParam = data.frame(m = (c*m + n*mean(data))/(c + n), 
                c = c + n, 
                a = a + n/2, 
                b =  b + 0.5*(sum((data - mean(data))^2)) + 
                  (n*c *(mean(data)- m)^2)/(2*(c+n)))
  return(postParam)
}
postS = findParam(prior, x)
postC = findParam(prior, y)

```
```{r, results = 'asis', echo = FALSE}
xtable(rbind(prior = prior,
             `Spurters Posterior` = postS,
             `Controls Posterior` = postC), 
       caption = "Parameters")
```

```{r}
# sampling from two posteriors 

# Number of posterior simulations
sim = 1000

# initialize vectors to store samples
mus = NULL
lambdas = NULL
muc = NULL
lambdac = NULL

# Following formula from the NormalGamma with 
# the update paramaters accounted accounted for below 

lambdas = rgamma(sim, shape = postS$a, rate = postS$b)
lambdac = rgamma(sim, shape = postC$a, rate = postC$b)


mus = sapply(sqrt(1/(postS$c*lambdas)),rnorm, n = 1, mean = postS$m)
muc = sapply(sqrt(1/(postC$c*lambdac)),rnorm, n = 1, mean = postC$m)

# Store simulations
simDF = data.frame(lambda = c(lambdas, lambdac),
                   mu = c(mus, muc),
                   Treatment = rep(c("Spurters", "Controls"),
                                   each = sim))

simDF$lambda = simDF$lambda^{-0.5}

# Plot the simulations
ggplot(data = simDF, aes(x = mu, y = lambda, colour = Treatment, shape = Treatment)) +
  geom_point(alpha = 0.2) + 
  labs(x = expression(paste(mu, " (Mean Change in IQ Score)")),
       y = expression(paste(lambda^{-1/2}, " (Std. Dev. of Change)")))  + 
  ggtitle("Posterior Samples")+ 
  theme(plot.title = element_text(hjust = 0.5))
```

Now, we can answer our original question: ``What is the posterior probability that $\mu_S>\mu_C$?'' 

The easiest way to do this is to take a bunch of samples from each of the posteriors, and see what fraction of times we have $\mu_S>\mu_C$. This is an example of a Monte Carlo approximation (much more to come on this in the future). 

To do this, we draw $N=10^6$ samples from each posterior:


```{r}
mp <- 24
cp <- 8
ap <- 4
bp <- 855

ms <- 11.8
cs <- 49
as <- 24.5
bs <- 6344.0

#Sampling from posterior
S <- 1000000


# Number of posterior simulations
sim = 1000000

# initialize vectors to store samples
mu_s = NULL
lambda_s = NULL
mu_c = NULL
lambda_c = NULL

# Following formula from the NormalGamma with 
# the update paramaters accounted accounted for below 

lambda_c = rgamma(S, as, bs)
lambda_s = rgamma(S, ap, bp)

mu_c = rnorm(S, ms, sd = sqrt(1/(cs*lambda_c)))
mu_s = rnorm(S, mp, sd = sqrt(1/(cp*lambda_s)))

# Store simulations
dataF = data.frame(lambda = c(lambda_s, lambda_c),
                   mu = c(mu_s, mu_c),
                   Treatment = rep(c("Spurters", "Controls"),
                                   each = sim))

dataF$lambda = dataF$lambda^{-0.5}
mean(mu_s > mu_c)
```
Interpret the posterior probability that you compute above. 

After sampling 10^6 times, we get 0.970831. 0.970831 means that 97.0831% of people who were told they were spurters(special group) had a higher change in IQ than the change in IQ of the control group. This also means that the probability that the spurters(special group) have a higher change in IQ than the change in IQ of the control group is 97.0831%.


# (b) (15) Task 5 (Finish for Homework)

Let's return back to the prior assumptions. There are a few ways that you can check that the prior conforms with our prior beliefs. Let's go back and checks these. Draw some samples from the prior and look at them---this is probably the best general strategy. See Figure pygmalion-prior. It's also a good idea to look at sample hypothetical datasets $X_{1:n}$ drawn using these sampled parameter values. 

Please replicate a plot similar to Figure pygmalion-prior and report your findings. 


